### 2025-04-09 correlation.qmd
20250408 - visualization and partial correlation

```{r}
library("petersenlab")
library("XICOR")
library("psych")
library("DescTools")
library("corrplot")
library("tidyverse")
```

@@

@@

### Partial Correlation {#sec-correlationExamplesNonlinear}

>>>>

```{r}
library("petersenlab")
library("XICOR")
library("psych")
library("DescTools")
library("psych")
library("corrplot")
library("tidyverse")
```

@@

### Scatterplot with Best-Fit Line

A scatterplot with best-fit line is in @fig-scatterplotWithBestFitLine.

```{r}
#| label: fig-scatterplotWithBestFitLine
#| fig-cap: "Scatterplot of Fantasy Points (Season) by Player Height With Best-Fit Line. The linear best-fit line is in black. The nonlinear best-fit line is in blue."
#| fig-alt: "Scatterplot of Fantasy Points (Season) by Player Height With Best-Fit Line. The linear best-fit line is in black. The nonlinear best-fit line is in blue."

plot_scatterplot <- ggplot2::ggplot(
  data = player_stats_seasonal %>% filter(position %in% c("QB","RB","WR","TE")),
  aes(
    x = height,
    y = fantasyPoints)) +
  geom_point(
    aes(
      text = player_display_name, # add player name for mouse over tooltip
      label = season), # add season for mouse over tooltip
    alpha = 0.05) +
  geom_smooth(
    method = "lm",
    color = "black") +
  geom_smooth() +
  coord_cartesian(
    ylim = c(0,NA),
    expand = FALSE) +
  labs(
    x = "Player Height (Inches)",
    y = "Fantasy Points (Season)",
    title = "Fantasy Points (Season) by Player Height",
    subtitle = "(Among QBs, RBs, WRs, and TEs)"
  ) +
  theme_classic()

plotly::ggplotly(plot_scatterplot)
```

Examining the linear line of best fit, we would see that there is a slight negative slope, consistent with a weak negative association.
However, the nonlinear best-fit line suggests that there may be an inverted-U-shaped association, which might suggest that there may be an "optimal range of height," where being too tall or too short may be a disadvantage.
We now examine the association between height and fantasy points separately by position in @fig-scatterplotByPositionWithBestFitLine.

```{r}
#| label: fig-scatterplotByPositionWithBestFitLine
#| fig-cap: "Scatterplot of Fantasy Points (Season) by Player Height and Position With Best-Fit Line. The linear best-fit line for each position is in black. The nonlinear best-fit line for each group is represented by the color line."
#| fig-alt: "Scatterplot of Fantasy Points (Season) by Player Height and Position With Best-Fit Line. The linear best-fit line for each position is in black. The nonlinear best-fit line for each group is represented by the color line."

plot_scatterplotByPosition <- ggplot2::ggplot(
  data = player_stats_seasonal %>% filter(position %in% c("QB","RB","WR","TE")),
  aes(
    x = height,
    y = fantasyPoints,
    color = position,
    fill = position)) +
  geom_point(
    aes(
      text = player_display_name, # add player name for mouse over tooltip
      label = season), # add season for mouse over tooltip
    alpha = 0.7) +
  geom_smooth(
    method = "lm",
    color = "black") +
  geom_smooth(
    method = "loess",
    span = 0.5) +
  coord_cartesian(
    ylim = c(0,NA),
    expand = FALSE) +
  labs(
    x = "Player Height (Inches)",
    y = "Fantasy Points (Season)",
    title = "Fantasy Points (Season) by Player Height and Position"
  ) +
  guides(fill = "none") +
  theme_classic()

plotly::ggplotly(plot_scatterplotByPosition)
```

Examining the association between height and fantasy points separately by position, it appears that there is an inverted-U-shaped associations for Wide Receivers, a relatively flat association for Running Backs and Tight Ends, and a nonlinear association among Quarterbacks.
Interestingly, the best-fit lines suggest that the different positions have different height ranges, which is confirmed in @fig-ridgelinePlotPlayerHeightByPosition.

```{r}
#| label: fig-ridgelinePlotPlayerHeightByPosition
#| fig-cap: "Ridgeline Plot of Player Height by Position."
#| fig-alt: "Ridgeline Plot of Player Height by Position."

ggplot2::ggplot(
  data = player_stats_seasonal %>%
    filter(position %in% c("QB","RB","WR","TE")),
  mapping = aes(
    x = height,
    y = position,
    group = position,
    fill = position)
) +
  ggridges::geom_density_ridges(
    bandwidth = 0.6
  ) +
  labs(
    x = "Height (inches)",
    y = "Position",
    title = "Ridgeline Plot of Player Height by Position"
  ) +
  theme_classic() + 
  theme(
    legend.position = "none",
    axis.title.y = element_text(angle = 0, vjust = 0.5)) # horizontal y-axis title
```

According to the distributions of player height by position (among Quarterbacks, Running Backs, Wide Receivers, and Tight Ends), Running Backs tend to be the shortest, followed by Wide Receivers; Tight Ends tend to be the tallest, followed by Quarterbacks.
Wide Receivers showed the greatest variability; some were quite short, whereas others were quite tall.

@@

### Partial Correlation {#sec-correlationExamplesPartial}

A partial correlation examines the association between variables after controlling for one or more variables.
Below, we examine a partial [Pearson correlation](#sec-correlationExamplesPearson) between height and fantasy points after controlling for age.

```{r}
psych::partial.r(
  data = player_stats_seasonal,
  x = c("height","fantasyPoints"),
  y = "age"
)
```

Below, we examine a partial [Spearman's rho rank correlation](#sec-correlationExamplesSpearman) between height and fantasy points after controlling for age.

```{r}
psych::partial.r(
  data = player_stats_seasonal,
  x = c("draftround","fantasyPoints"),
  y = "age",
  method = "spearman"
)
```
